---
title: "Quantifying macro-evolutionary patterns of trait mean and variance with phylogenetic location-scale models"
author: "**Shinichi Nakagawa, Ayumi Mizuno, Coralie Williams, Malgorzata Lagisz, Yefeng Yang and Szymon M Drobniak**"
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 3
    toc-title: "**Table of Contents**"
    output-file: "index.html"
    theme: simplex
    embed-resources: true
    code-fold: show
    code-tools: true
    number-sections: true
    #bibliography: ./bib/ref.bib
    fontsize: "12"
    max-width: "10"
    code-overflow: wrap
crossref: 
  fig-title: Figure     # (default is "Figure")
  tbl-title: Table     # (default is "Table")
  title-delim: —     # (default is ":")
  fig-prefix: Fig.   # (default is "Figure")
  tbl-prefix: Tab.    # (default is "Table")
editor_options: 
  chunk_output_type: console
editor: 
  markdown: 
    wrap: sentence
---

```{r setup}
#| include: false

knitr::opts_chunk$set(
  collapse = TRUE,
  message = FALSE,
  warnings = FALSE,
  echo = TRUE#,
  #comment = "#>"
)
```

# Update

Last update March 2025.

We will update this tutorial when necessary.
Readers can access the latest version in our [GitHub repository](https://github.com/itchyshin/phylo_location_scale).

If you have any questions, errors or bug reports, please contact Szymon M Drobniak (szymek.drobniak\@gmail.com) or Shinichi Nakagawa (snakagaw\@ualberta.ca).

# Introduction

This online material is a supplement to our paper "Quantifying macro-evolutionary patterns of trait mean and variance with phylogenetic location-scale models".
You will see how to fit phylogenetic location-scale models (PLSMs) in a Bayesian framework using `brms` package in `R`.

# Contents

In this online material, we will illustrate how to fit three models described in our paper, subtitled as:

- Different Trait Variance in Two Groups (Model 3)

- Co-evolution of Mean and Variance (Model 4)

- Co-evolution of Two Traits (Model 5)

All results presented in this online material are for illustrative purposes only (more careful thoughts are required to draw biological conclusions). 

Additionally, we will show how to obtain phylogenetic heritablity and evolvability from the fitted `brms` object.

Also, importantly, there is a sister paper to this one, which focuses on locatoin-scale meta-analytic and meta-regression. You can find the sister paper's tutorial [here](https://itchyshin.github.io/location-scale_meta-analysis/#multilevel-meta-analysis-and-location-scale-meta-regression); also, the preprint of the sister paper is available on [EcoEvoRxiv](https://ecoevorxiv.org/repository/view/8540/):

> Nakagawa S, Mizuno A, Morrison K, Ricolfi L, Williams C, Drobniak SM, Lagisz M, Yang Y. Location-scale Meta-analysis and Meta-regression as a Tool to Capture Large-scale Changes in Biological and Methodological Heterogeneity: a Spotlight on Heteroscedasticity. EcoEvoRxiv. 2025. 

This tutorial has several tips to run location-scale models using `brms`. 

# Prerequisites

## Loading pacakges

Our tutorial uses `R` statistical software and existing `R` packages, which you will first need to download and install.

If the packages are archived in CRAN, use `install.packages()` to install them.
For example, to install the `brms` , you can execute `install.packages("brms")` in the console (bottom left pane of `R Studio`).

Version information of each package is listed at the end of this tutorial.

```{r packages}
# attempt to load or install necessary packages
if(!require(pacman)) install.packages("pacman")
pacman::p_load(tidyverse, 
               ggplot2,
               brms,
               bayesplot,
               tidybayes,
               here,
               ape,
               patchwork,
               pander
)

```

## Custom functions

We also provide some additional helper functions to help visualize the results (mainly extracted from the `brms` object).
If you want to use these custom functions in your own data, you'll need to change the variable names according to your own data (check out the `R` code and you'll see what we mean).

```{r}
#| code-fold: true

# Function to get variable names dynamically
get_variables_dynamic <- function(model, pattern) {
  variables <- get_variables(model)
  variables[grep(pattern, variables)]
}

rename_vars <- function(variable) {
  # b
  variable <- gsub("b_Intercept", "l_int", variable)
  variable <- gsub("b_sigma_Intercept", "s_int", variable)  
  variable <- gsub("b_cmass", "l_cmass", variable)            
  variable <- gsub("b_sigma_cmass", "s_cmass", variable)
  variable <- gsub("b_forest", "l_contrast", variable)            
  variable <- gsub("b_sigma_forest", "s_contrast", variable) 
  # mod3
  variable <- gsub("b_cbeakwidth_Intercept", "l_width_int", variable)
  variable <- gsub("b_sigma_cbeakwidth_Intercept", "s_width_int", variable)  
  variable <- gsub("b_cbeakdepth_Intercept", "l_depth_int", variable)            
  variable <- gsub("b_sigma_cbeakdepth_Intercept", "s_depth_int", variable)
  variable <- gsub("b_cbeakwidth_cmass", "l_width_cmass", variable)            
  variable <- gsub("b_sigma_cbeakwidth_cmass", "s_width_cmass", variable) 
  variable <- gsub("b_cbeakdepth_cmass", "l_depth_cmass", variable)            
  variable <- gsub("b_sigma_cbeakdepth_cmass", "s_depth_cmass", variable) 
  
  # sd
  variable <- gsub("sd_Phylo__Intercept", "l_sd", variable)  
  variable <- gsub("sd_Phylo__sigma_Intercept", "s_sd", variable)  
  # mod3
  variable <- gsub("sd_Phylo__cbeakwidth_Intercept", "l_width_sd", variable)  
  variable <- gsub("sd_Phylo__sigma_cbeakwidth_Intercept", "s_width_sd", variable)  
  variable <- gsub("sd_Phylo__cbeakdepth_Intercept", "l_depth_sd", variable)  
  variable <- gsub("sd_Phylo__sigma_cbeakdepth_Intercept", "s_depth_sd", variable) 
  
  # corr
  variable <- gsub("cor_Phylo__Intercept__sigma_Intercept", "ls_cor", variable)  
  # mod3
  variable <- gsub("cor_Phylo__cbeakwidth_Intercept__sigma_cbeakwidth_Intercept", "ls_width-width_cor", variable) 
  variable <- gsub("cor_Phylo__cbeakwidth_Intercept__cbeakdepth_Intercept", "ll_width-depth_cor", variable) 
  variable <- gsub("cor_Phylo__sigma_cbeakwidth_Intercept__cbeakdepth_Intercept", "sl_width-depth_cor", variable) 
  variable <- gsub("cor_Phylo__cbeakwidth_Intercept__sigma_cbeakdepth_Intercept", "ls_width-depth_cor", variable) 
  variable <- gsub("cor_Phylo__sigma_cbeakwidth_Intercept__sigma_cbeakdepth_Intercept", "ss_width-depth_cor", variable) 
  variable <- gsub("cor_Phylo__cbeakdepth_Intercept__sigma_cbeakdepth_Intercept", "ls_depth-depth_cor", variable) 
  
  return(variable)
}


# Function to visualize fixed effects
visualize_fixed_effects <- function(model) {
  fixed_effect_vars <- get_variables_dynamic(model, "^b_")
  if (length(fixed_effect_vars) == 0) {
    message("No fixed effects found")
    return(NULL)
  }
  
  tryCatch({
    fixed_effects_samples <- model %>%
      spread_draws(!!!syms(fixed_effect_vars)) %>%
      pivot_longer(cols = all_of(fixed_effect_vars), names_to = ".variable", values_to = ".value") %>%
      mutate(.variable = rename_vars(.variable))
    
    ggplot(fixed_effects_samples, aes(x = .value, y = .variable)) +
      stat_halfeye(
        normalize = "xy", 
        point_interval = "mean_qi", 
        fill = "lightcyan3", 
        color = "lightcyan4"
      ) +
      geom_vline(xintercept = 0, linetype = "dashed", color = "#005") +
      labs(y = "Fixed effects", x = "Posterior values") +
      theme_classic()
  }, error = function(e) {
    message("Error in visualize_fixed_effects: ", e$message)
    return(NULL)
  })
}

# Function to visualize random effects
visualize_random_effects <- function(model) {
  random_effect_vars <- get_variables_dynamic(model, "^sd_")
  random_effect_vars <- random_effect_vars[random_effect_vars != "sd_es_ID__Intercept"]
  if (length(random_effect_vars) == 0) {
    message("No random effects found")
    return(NULL)
  }
  
  tryCatch({
    random_effects_samples <- model %>%
      spread_draws(!!!syms(random_effect_vars)) %>%
      pivot_longer(cols = all_of(random_effect_vars), names_to = ".variable", values_to = ".value") %>%
      mutate(.variable = rename_vars(.variable)) #%>%
      #mutate(.value = .value)  # leave SD as it is
    
    ggplot(random_effects_samples, aes(x = .value, y = .variable)) +
      stat_halfeye(
        normalize = "xy", 
        point_interval = "mean_qi", 
        fill = "olivedrab3", 
        color = "olivedrab4"
      ) +
      geom_vline(xintercept = 0, linetype = "dashed", color = "#005") +
      labs(y = "Random effects (SD)", x = "Posterior values") +
      theme_classic()
  }, error = function(e) {
    message("Error in visualize_random_effects: ", e$message)
    return(NULL)
  })
}

# Function to visualize correlations
visualize_correlations <- function(model) {
  correlation_vars <- get_variables_dynamic(model, "^cor_")
  if (length(correlation_vars) == 0) {
    message("No correlations found")
    return(NULL)
  }
  
  tryCatch({
    correlation_samples <- model %>%
      spread_draws(!!!syms(correlation_vars)) %>%
      pivot_longer(cols = all_of(correlation_vars), names_to = ".variable", values_to = ".value") %>%
      mutate(.variable = rename_vars(.variable))
    
    ggplot(correlation_samples, aes(x = .value, y = .variable)) +
      stat_halfeye(
        normalize = "xy", 
        fill = "#FF6347", 
        color = "#8B3626"
      ) +
      geom_vline(xintercept = 0, linetype = "dashed", color = "#005") +
      labs(y = "Correlations", x = "Posterior values") +
      theme_classic()
  }, error = function(e) {
    message("Error in visualize_correlations: ", e$message)
    return(NULL)
  })
}
```


# Data loding and preparation

We use the data from **Avonet**:

> J. A. Tobias, C. Sheard, A. L. Pigot, A. J. Devenish, J. Yang, F. Sayol, M. H. Neate-Clegg, N. Alioravainen, T. L. Weeks, R. A. Barber, et al.
> Avonet: morphological, ecological and geographical data for all birds.
> Ecology Letters, 25(3):581–597, 2022

As described in the paper, we only use parrows species (*N* = 354) and we use phylogenetic trees form Jetz et al. (2012):

> W. Jetz, G. H. Thomas, J. B. Joy, K. Hartmann, and A. O. Mooers. The global diversity of birds in space and time.
Nature, 491(7424):444–448, 2012

```{r}
# load data
dat <- read.csv(here("data", "Psittaciformes", "Psittaciformes_354spp.csv"))

#str(dat)

# reading nexus file (use ape pacakge)
tree_all <- read.nexus(here("data", "Psittaciformes", "Psittaciformes_354spp_100.nex"))
tree <- tree_all[[1]]

# Turning phylogenetic tree into phylognetic correlation matrix
A <- vcv.phylo(tree, corr = TRUE)

# center data of interest
dat$cbeak_length <- scale(log(dat$Beak.Length_Culmen), center = TRUE, scale = FALSE)
dat$cbeak_width <- scale(log(dat$Beak.Width), center = TRUE, scale = FALSE)
dat$cbeak_depth <- scale(log(dat$Beak.Depth), center = TRUE, scale = FALSE)
dat$cmass <- scale(log(dat$Mass), center = TRUE, scale = FALSE)
dat$crange_size <- scale(log(dat$Range.Size), center = TRUE, scale = FALSE)

# phynotipic correlation among three beak traits
# cor(dat$cbeak_length, dat$cbeak_width)
# cor(dat$cbeak_length, dat$cbeak_depth)
# cor(dat$cbeak_width, dat$cbeak_depth)

# create a variable which says whether they live in Forest or not
dat$forest <- ifelse(dat$Habitat == "Forest", 1, 0)
```

We use the following variables in our models:

- `cbeak_length`: log-transformed and centred beak length (`Beak.Length_Culmen`)

- `cbeak_width`: log-transformed and centred beak width (`Beak.Width`)

- `cbeak_depth`: log-transformed and centred beak depth (`Beak.Depth`)

- `cmass`: log-transformed and centred body mass (`Mass`)

- `crange_size`: log-transformed and centred range size (`Range.Size`)

- `forest`: a binary variable indicating whether the species lives in the forest or other habits (created based on `Habitat`)


# Different trait variance in two groups (Model 3)

In this first model, we will fit a location-scale model to investigate whether the mean and variance of beak length (`cbeak_length`) differs between species living in the forest and those living in other habitats.

## Fitting the model

The `brms` syntax for Model 3 is as follows:

```{r}
#| eval: false

# creating the formula
form1 <- bf(cbeak_length ~ 1 + cmass + forest + (1 | gr(Phylo, cov = A)) 
         sigma ~ 1 + cmass + forest
)

# setting priors
prior1 <- default_prior(form1, 
                        data = dat, 
                        data2 = list(A = A),
                        family = gaussian()
)

# running the model
mod1 <- brm(form1, 
            data = dat, 
            data2 = list(A = A),
            chains = 2, 
            cores = 2, 
            iter = 40000, 
            warmup = 10000,
            prior = prior1,
            control = list(adapt_delta = 0.99, max_treedepth = 15)
)

# saving
saveRDS(mod1, here("Rdata", "mod1.rds"))
```

Check the results: 

```{r}
# loading and checking
mod1 <- readRDS(here("Rdata", "mod1.rds"))

summary(mod1)
```

## Visualizing results

We can visualize the results (posteior distributions) of parameters (i.e., regression coefficents, variance/standard deviation components and correlations) using the custom functions we defined earlier.

```{r}
#| warning: false
#| fig-height: 6

# getting plots
plots_mod1 <- list(
  visualize_fixed_effects(mod1),
  visualize_random_effects(mod1)
)

plots_mod1[[1]] / plots_mod1[[2]] + plot_layout(heights = c(2, 1))
```

**Posterior distributions of different types of parameters**; the vertical dashed line indicates zero, aiding the interpretation of effect direction with thick lines showing 66% credible intervals and thin whiskers 95% credible intervals (not spanning zero being interpreted as 'statistical significant'). 

# Co-evolution of mean and variance (Model 4)

In this second model, we will fit a phylogenetic location-scale model to investigate whether the mean and variance of beak length (`cbeak_length`) are evolving together among parrots, after controlling body mass (`cmass`).

## Fitting the model

The `brms` syntax for Model 4 is as follows:

```{r}
#| eval: false

# creating the formula
# |p| indicates we are going to estimate correlations between these two random effects
form2 <- bf(crange_size ~1 + cmass + (1|p|gr(Phylo, cov = A)), 
            sigma ~ 1 + cmass + (1|p|gr(Phylo, cov = A))
)

# setting priors
prior2 <- default_prior(form2, 
                        data = dat, 
                        data2 = list(A = A),
                        family = gaussian()
)

# running the model
mod2 <- brm(form2, 
            data = dat, 
            data2 = list(A = A),
            chains = 2, 
            cores = 2, 
            iter = 6000, 
            warmup = 3000,
            prior = prior2,
            control = list(adapt_delta = 0.99, max_treedepth = 15)
)

# saving
saveRDS(mod2, here("Rdata", "mod2.rds"))

```

Check the results: 

```{r}
# loading and checking
mod2 <- readRDS(here("Rdata", "mod2.rds"))

summary(mod2)

```


## Visualizing results

As above, using the custom functions:

```{r}
#| warning: false
#| fig-height: 8

# getting plots
plots_mod2 <- list(
  visualize_fixed_effects(mod2),
  visualize_random_effects(mod2),
  visualize_correlations(mod2)
)


plots_mod2[[1]] / plots_mod2[[2]] /plots_mod2[[3]] + plot_layout(heights = c(2, 1, 1))
```

**Posterior distributions of different types of parameters**; the vertical dashed line indicates zero, aiding the interpretation of effect direction with thick lines showing 66% credible intervals and thin whiskers 95% credible intervals (not spanning zero being interpreted as 'statistical significant'). 

# Co-evolution of two traits (Model 5)

In this third model, we will fit a phylogenetic bivariate location-scale model to investigate whether the mean and variance of beak length (`cbeak_length`) and beak width (`cbeak_width`) are evolving together among parrots, after controlling body mass (`cmass`).

## Fitting the model

The `brms` syntax for Model 4 is as follows:
  
```{r}
#| eval: false

# creating the formula
# |p| indicates we are going to estimate correlations between these 4 random effects
form3A <- bf(cbeak_width ~1 + cmass + (1|p|gr(Phylo, cov = A)), 
                sigma ~ 1 + cmass + (1|p|gr(Phylo, cov = A))
)

form3B <- bf(cbeak_depth ~1 + cmass + (1|p|gr(Phylo, cov = A)), 
                sigma ~ 1 + cmass + (1|p|gr(Phylo, cov = A))
)

form3 <- form3A + form3B + set_rescor(TRUE) 

# setting prior
prior3 <- default_prior(form3, 
                        data = dat, 
                        data2 = list(A = A),
                        family = gaussian()
)

# fit model
mod3 <- brm(form3, 
                  data = dat, 
                  data2 = list(A = A),
                  chains = 2, 
                  cores = 2, 
                  iter = 15000, 
                  warmup = 5000,
                  prior = prior3,
                  control = list(adapt_delta = 0.99, max_treedepth = 15)
)

summary(mod3)

# saving
saveRDS(mod3, here("Rdata", "mod3.rds"))
```

Check the results: 

```{r}
# loading and checking
mod3 <- readRDS(here("Rdata", "mod3.rds"))

summary(mod3)
```


## Visualizing results

Again, using the custom functions:

```{r}
#| warning: false
#| fig-height: 10

# getting plots
plots_mod3 <- list(
  visualize_fixed_effects(mod3),
  visualize_random_effects(mod3),
  visualize_correlations(mod3)
)


plots_mod3[[1]] / plots_mod3[[2]] /plots_mod3[[3]]
```

**Posterior distributions of different types of parameters**; the vertical dashed line indicates zero, aiding the interpretation of effect direction with thick lines showing 66% credible intervals and thin whiskers 95% credible intervals (not spanning zero being interpreted as 'statistical significant'). 

# Bonus: obtaining phylogenetic heritability and evolvability (phylogenetic CV)

Below, we show how to calculate phylogenetic heritability and evolvability from the fitted `brms` object with and without a fixed effect(s). As you will see, phylogenetic hertiaiblity is fairly straightforward to obtain but evolvability is a bit more complex as it is not clear what sacle to calculate it on (e.g., log-scale or original scale). In fact, the same issue should be relevant to phylogenetic heritability as well but we do not consider this here. 

## a model without fixed effects

We first create a phylogenetic location-scale (double-hierarchical) model for `cmass` without any fixed effects. 

The `brms` syntax for this phylogenetic location-scale model is as follows:

```{r}
#| eval: false

# creating the formula
# |p| indicates we are going to estimate correlations between these 2 random effects
form0 <- bf(crange_size ~ 1  + (1 | p | gr(Phylo, cov = A)),
            sigma ~ 1 + (1 | p | gr(Phylo, cov = A))
)

# setting prior
prior0 <- default_prior(form0, 
                        data = dat, 
                        data2 = list(A = A),
                        family = gaussian()
)

# fit model
mod0 <- brm(form0, 
            data = dat, 
            data2 = list(A = A),
            chains = 2, 
            cores = 2, 
            iter = 8000, 
            warmup = 1000,
            prior = prior0,
            control = list(adapt_delta = 0.99, max_treedepth = 15)
)

# saving
saveRDS(mod0, here("Rdata", "mod0.rds"))
```

Check the results: 

```{r}
# loading and checking
mod0 <- readRDS(here("Rdata", "mod0.rds"))

summary(mod0)
```


### Phylogenetic heritability

Using Equations 14,  15 & 17 in the paper, we can calculate the phylogenetic heritability of the trait `crange_size` (the location part):

```{r}

# sigma2_p (total phenotypic variance)

# posterior draws
post <- as_draws_df(mod0)

# Equation 17: first calculate sigma2_e_bar (average residual variance)
sigma2_e_bar <- exp(2*post$b_sigma_Intercept + 2*post$sd_Phylo__sigma_Intercept^2)

# Equation 15 (no fixed effects)
sigma2_p <- post$sd_Phylo__Intercept^2 +  sigma2_e_bar

# Equation 14

H2l <- post$sd_Phylo__Intercept^2 / sigma2_p

#summary(H2)
quantile(H2l, probs = c(0.025, 0.5, 0.975)) %>% pander()
```

We can see the median phylogenetic heritablity is around 38%.

Using Equations 18-21 in the paper, we can calculate the phylogenetic heritability of the trait `crange_size` (the scale part):

```{r}
# Eqatuion 21 - sigma2_sigma2_e (variance of the residual variance)
sigma2_sigma2_e <- (exp(4*post$sd_Phylo__sigma_Intercept^2) - 1) * exp(4*(post$b_sigma_Intercept + post$sd_Phylo__sigma_Intercept^2))

# Equation 19 - sigma2_a_star 
sigma2_as_star <- sigma2_sigma2_e 

# Equation 20 - sigma2_sigma2_p (variance of the total phenotypic variance)
sigma2_sigma2_p = sigma2_p ^ 2 + 3 * sigma2_sigma2_e

# Equation 18
H2s <- sigma2_as_star / sigma2_sigma2_p

#summary(H2)
quantile(H2s, probs = c(0.025, 0.5, 0.975)) %>% pander()
```

Here we found a median phylogenetic heritability of around 31%.

### Evolvability

Using Equations 6 in the paper, we can calculate the evolvability (CV_Al) of the trait `crange_size` (log-transformed trait; the location part), but the difficulty is that we need to calculate the phylogenetic variance on the orginal scale. We can do this by using the delta method. The details of this calculation are provided in this earlier work (see the supplmentary material):

> O'Dea RE, Noble DW, Nakagawa S. Unifying individual differences in personality, predictability and plasticity: a practical guide. Methods in Ecology and Evolution. 2022 Feb;13(2):278-93.

```{r}
# exp(b0) is the median of the trait
ln_b0 <- post$b_Intercept  + mean(log(dat$Range.Size), na.rm = T)
sigma2_al <- post$sd_Phylo__Intercept^2 

mean_b0 <- exp(ln_b0 +  0.5 * (sigma2_al +  sigma2_e_bar) )

# conversion formula from log-scale to original scale
sd_al <- sqrt( (exp(sigma2_al) - 1) * exp(2*ln_b0 + sigma2_al) )

CV_Al <- sd_al / mean_b0

quantile(CV_Al, probs = c(0.025, 0.5, 0.975)) %>% pander()
```

We can see the median evolvability is around 0.08

```{r}
# alternatively, we can get CV on ln-scale
CV_Al2 <- sigma2_al / ln_b0

quantile(CV_Al2, probs = c(0.025, 0.5, 0.975)) %>% pander()
```

If we get the CV on the log-scale, we can see the median evolvability is around 0.85 (nearly 10 times bigger); this CV for the location is probably more comparable to CV for the scale below (as it turns, our CV for the scale is not estimable on the original scale). 

We now estimate the evolvability of the trait `crange_size` (the scale part) using Equatoin 24 (we assume that this CV_As is the same as the value for the original scale):

```{r}

CV_As <- sqrt(exp(4*post$sd_Phylo__sigma_Intercept^2) - 1)

quantile(CV_As, probs = c(0.025, 0.5, 0.975)) %>% pander()
```

We get the median evolvability of around 3.4. This is evolvability of variance but we can get evolvability of the standard deviation using Equation 25. 

```{r}
# Alternatively, we can get CV_As on the SD scale

CV_As2 <- sqrt(exp(post$sd_Phylo__sigma_Intercept^2) - 1)
quantile(CV_As2, probs = c(0.025, 0.5, 0.975)) %>% pander()
```

We get the median evolvability of around 0.94, which is not that different from the evolvability of the mean estiamted on the log scale. 

Also, we can use Equation 23:

```{r}
CV_As2 <- sqrt(sigma2_as_star) / sigma2_e_bar

quantile(CV_As2, probs = c(0.025, 0.5, 0.975), na.rm = T) %>% pander()
```

This estimation, as expected, matches with that from Equation 24. 

We also try to get the evolvability for the sacle on the original scale

```{r}
# original scale conversion does not seem to make sense so commented out
sigma2_as_star_orig <- sqrt( (exp(sigma2_as_star) - 1) * exp(2*ln_b0 + sigma2_as_star) )
sigma2_e_bar_orig <- sqrt( (exp(sigma2_e_bar) - 1) * exp(2*ln_b0 + sigma2_e_bar) )

CV_As2_orig <- sqrt(sigma2_as_star_orig) / sigma2_e_bar_orig

quantile(CV_As2_orig, probs = c(0.025, 0.5, 0.975), na.rm = T) %>% pander()
```

We get the median evolvability of a `Inf` value so this does not seem to work. 

## a model with fixed effects

Here we use `mod2` as an example to see the hertiablity and evolvability of the trait `crange_size` after controlling for body mass `cmass`.

### Phylogenetic heritability

Using Equations 14-17 in the paper, we can calculate the phylogenetic heritability of the trait `crange_size` (the location part) after controlling for body mass `cmass`:

```{r}

# sigma2_p (total phenotypic variance)
# posterior draws
post <- as_draws_df(mod2)

# Equation 17: first calculate sigma2_e_bar (average residual variance)
sigma2_e_bar <- exp(2*post$b_sigma_Intercept + 2*post$sd_Phylo__sigma_Intercept^2)


# Equation 16 - sigma2_f (variance of the fixed effects)
sigma2_fl <- sapply(post$b_cmass, function(b) var(b*dat$cmass))

# Equation 15 (no fixed effects)
sigma2_p <- post$sd_Phylo__Intercept^2 + sigma2_fl  + sigma2_e_bar

# Equation 14

H2l <- post$sd_Phylo__Intercept^2 / sigma2_p

#summary(H2)
quantile(H2l, probs = c(0.025, 0.5, 0.975)) %>% pander()
```

We can see the median phylogenetic heritablity is around 45%.

Using Equations 18-21 in the paper, we can calculate the phylogenetic heritability of the trait `crange_size` (the scale part):

```{r}

# Equation 22 - sigma2_fs
sigma2_fs <- sapply(post$b_sigma_cmass, function(b) var(b*dat$cmass))

# Equation 21 - sigma2_sigma2_e (variance of the residual variance)
sigma2_sigma2_e <- (exp(4*(post$sd_Phylo__sigma_Intercept^2 + sigma2_fs)) - 1) * exp(4*(post$b_sigma_Intercept + post$sd_Phylo__sigma_Intercept^2 + sigma2_fs))

# Equation 19 - sigma2_a_star 
sigma2_as_star <- sigma2_sigma2_e * (post$sd_Phylo__sigma_Intercept^2 /(post$sd_Phylo__sigma_Intercept^2 + sigma2_fs))

# Equation 20 - sigma2_sigma2_p (variance of the total phenotypic variance)
sigma2_sigma2_p = sigma2_p ^ 2 + 3 * sigma2_sigma2_e

# Equation 18
H2s <- sigma2_as_star / sigma2_sigma2_p

#summary(H2)
quantile(H2s, probs = c(0.025, 0.5, 0.975)) %>% pander()

```

Here we found a median phylogenetic heritability of around 28%.

### Evolvability

Using Equations 6 in the paper, we can calculate the evolvability (CV_Al) of the trait `crange_size` (log-transformed trait; the location part), but we first need to back-trasform like we did above. 

```{r}

# exp(b0) is the median of the trait
ln_b0 <- post$b_Intercept  + mean(log(dat$Range.Size), na.rm = T)
sigma2_al <- post$sd_Phylo__Intercept^2 

mean_b0 <- exp(ln_b0 +  0.5 * (sigma2_al +  sigma2_e_bar) )

# (exp(sigma_ln^2) - 1) * exp(2 * mu_ln + sigma_ln^2)
sd_al <- sqrt( (exp(sigma2_al) - 1) * exp(2*ln_b0 + sigma2_al) )

CV_Al <- sd_al / mean_b0


quantile(CV_Al, probs = c(0.025, 0.5, 0.975)) %>% pander()
```

We can see the median evolvability is around 0.38.

```{r}
# alternatively, we can get CV on ln-scale
CV_Al2 <- sigma2_al / ln_b0
quantile(CV_Al2, probs = c(0.025, 0.5, 0.975)) %>% pander()
```

If we get the CV on the log-scale, we can see the median evolvability is around 0.83, which is higher than what we got on the original scale (like above).

We now estimate the evolvability of the trait `crange_size` (the scale part) using Equatoin 24 (we assume that this CV_As is the same as the value for the original scale):

```{r}
# Using Equation 24
CV_As <- sqrt(exp(4*post$sd_Phylo__sigma_Intercept^2) - 1)

quantile(CV_As, probs = c(0.025, 0.5, 0.975)) %>% pander()
```

We get the median evolvability of around 3.2. As above, this is evolvability of variance but we can get evolvability of the standard deviation using Equation 25.

```{r}
# Alternatively, we can get CV_As on the SD scale
CV_As2 <- sqrt(exp(post$sd_Phylo__sigma_Intercept^2) - 1)

quantile(CV_As2, probs = c(0.025, 0.5, 0.975)) %>% pander()
```

We get the median evolvability of around 0.91, which is not that different from the evolvability of the mean estiamted on the log scale.

Additionally, we can use Equation 23:

```{r}
CV_As <- sqrt(sigma2_as_star) / sigma2_e_bar

quantile(CV_As, probs = c(0.025, 0.5, 0.975), na.rm = T) %>% pander()
```

This estimation does matches with that from Equation 24 unlike the above. This is because now we have a fixed effect which is taken into account in account for obtaining the phylogenetic variance on the scale part: `sigma2_as_star` (we will not try to get the evolvability on the original scale as it does not seem to work).

# Figure code

```{r}
#| code-fold: true
#| fig-height: 10
#| 
# putting all plots together 
(plots_mod1[[1]] + plots_mod1[[2]]) + 
(plots_mod2[[1]] / plots_mod2[[2]] /plots_mod2[[3]] ) + 
(plots_mod3[[1]] / plots_mod3[[2]] /plots_mod3[[3]]) + plot_layout(heights = c(1, 3), widths = c(1, 1)) + 
  # put A, B, C
  plot_annotation(tag_levels = 'A')
```

# Software and package versions

```{r}
#| code-fold: true
sessionInfo() %>% pander()
```
